---
title: "Transnational Family Migration and Loneliness Among Older Filipinos"
subtitle: "First-Difference + IPW Analysis with Mechanism and Exploratory Moderation"
author: "Jem Marie Nario"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    theme: cosmo
    code-fold: false
    code-tools: true
    self-contained: true
knitr:
  opts_chunk:
    warning: false
    message: false
    echo: true
editor: visual
---

# Overview and Analytic Strategy

This document presents the full analysis for **"Transnational Family Migration and Loneliness Among Older Filipinos"** using the Longitudinal Study of Aging and Health in the Philippines (LSAHP), Waves 1 and 2.

## Three-Tier Analytic Structure

The analysis is organised into three tiers reflecting different levels of causal identification and available sample size.

**Tier 1 — Causal (H1, n = 3,291).** A survey-weighted first-difference (FD) model with inverse probability weighting (IPW) estimates the within-person effect of entering a transnational household on loneliness. FD removes all time-invariant confounding; IPW additionally balances 15 observed baseline characteristics between enterers and non-enterers. The primary outcome is the continuous 3-item loneliness scale change (d_lonely3); the binary CES-D frequent loneliness item (d_lonely) serves as a robustness check.

**Tier 2 — Associational / Mechanism (H2, n = 3,075).** Among respondents with non-co-resident children, changes in contact satisfaction and contact frequency are added to the FD model. These estimates identify within-person associations between communication quality and loneliness change, net of transnational transition. Reverse causality (loneliness reducing satisfaction) cannot be fully ruled out; findings are interpreted as evidence of mechanism rather than strict causal effects.

**Tier 3 — Exploratory (H3 + H4, n = 395).** Distance models are restricted to respondents with recorded destination countries in both waves. Crucially, this subsample consists **entirely of households that were already transnational at Wave 1 and remained so at Wave 2** — there are zero enterers in this subsample. The distance analysis therefore does not identify the entry effect; instead it asks: *among already-transnational households, does a child moving farther away increase parental loneliness, and does contact satisfaction buffer this?* This is a distinct and meaningful question, but must be clearly distinguished from the Tier 1 causal claim.

## Key Missingness Facts

| Variable               | Missing | \%  |
|------------------------|---------|-----|
| d_lonely (binary)      | 2,648   | 44% |
| d_lonely3 (continuous) | 2,694   | 45% |
| d_contact_sat          | 2,869   | 48% |
| d_dist_1000km          | 5,513   | 92% |
| d_poor / d_urban       | 1,974   | 33% |
| d_health               | 2,692   | 45% |

The 92% missingness on distance is structural: destination country codes are only recorded when a child is abroad at both waves with a valid country code entered by the interviewer. The distance subsample is highly selected and findings must be treated as preliminary.

------------------------------------------------------------------------

# Setup

```{r}
#| label: setup
#| include: false

pkgs <- c(
  "dplyr", "tidyr", "ggplot2", "scales",
  "janitor", "haven", "forcats",
  "broom", "survey",
  "WeightIt", "cobalt",
  "marginaleffects", "modelsummary",
  "geosphere", "rnaturalearth", "sf",
  "DiagrammeR", "DiagrammeRsvg", "rsvg",
  "writexl", "knitr", "kableExtra",
  "patchwork"
)

install_if_missing <- pkgs[!pkgs %in% installed.packages()[, "Package"]]
if (length(install_if_missing) > 0) {
  install.packages(install_if_missing, repos = "https://cran.r-project.org")
}

invisible(lapply(pkgs, library, character.only = TRUE))

LMUGreen <- "#00883A"

theme_lmu <- function(base_size = 13) {
  theme_minimal(base_size = base_size) +
    theme(
      plot.title       = element_text(face = "bold"),
      plot.subtitle    = element_text(color = "grey40", size = 10),
      axis.title       = element_text(face = "bold"),
      panel.grid.minor = element_blank(),
      legend.position  = "bottom"
    )
}

save_plot <- function(p, name, w = 10, h = 6) {
  dir.create("figures", showWarnings = FALSE)
  ggsave(file.path("figures", paste0(name, ".pdf")),
         p, width = w, height = h, device = cairo_pdf)
  ggsave(file.path("figures", paste0(name, ".png")),
         p, width = w, height = h, dpi = 300)
  invisible(p)
}

dir.create("tables", showWarnings = FALSE)
```

------------------------------------------------------------------------

# Load Data

```{r}
#| label: load-data

# Remove only data objects, not helper functions
rm(list = ls()[!ls() %in% c("LMUGreen", "theme_lmu", "save_plot")])

library(haven)

lsahp1 <- read_dta(
  "C:/Users/jemma/Downloads/LSAHP-Baseline-Data-STATA/LSAHP Baseline Data (STATA)/LSAHP1.dta"
)

lsahp2 <- read_dta(
  "H:/My Drive/UPPI_RUG/UPPI_RUG_DLSU/LSAHP W2 Transnational households and loneliness (Jem, 202602)/LSAHP W2 Transnational households and loneliness (Jem).dta"
)

cat("Wave 1 n =", nrow(lsahp1), "\n")
cat("Wave 2 n =", nrow(lsahp2), "\n")
```

------------------------------------------------------------------------

# Wave 1 Variables

## ID

```{r}
#| label: w1-id
lsahp_w1 <- lsahp1 |> dplyr::mutate(id = P1ID1)
```

## Outcomes

```{r}
#| label: w1-outcomes

# Binary: 1 = often lonely (frequent loneliness threshold)
lsahp_w1 <- lsahp_w1 |>
  dplyr::mutate(
    lonely_w1 = dplyr::case_when(
      P1SD1_06 == 3          ~ 1,
      P1SD1_06 %in% c(1, 2) ~ 0,
      TRUE                   ~ NA_real_
    )
  )

# 3-item continuous scale (0-4): mean of companionship, left out, isolated
# Wave 1 uses same 0-4 response categories as Wave 2 — confirmed comparable
lsahp_w1 <- lsahp_w1 |>
  dplyr::mutate(
    lonely_3item = dplyr::if_else(
      rowSums(dplyr::across(c(P1SH9_1, P1SH9_2, P1SH9_3),
                            ~ is.na(.x))) == 0,
      rowMeans(dplyr::across(c(P1SH9_1, P1SH9_2, P1SH9_3))),
      NA_real_
    )
  )

cat("Wave 1 lonely_w1:\n")
table(lsahp_w1$lonely_w1, useNA = "ifany")
cat("\nWave 1 lonely_3item summary:\n")
summary(lsahp_w1$lonely_3item)
```

## Exposure: Transnational Household

```{r}
#| label: w1-transnational

lsahp_w1 <- lsahp_w1 |>
  dplyr::mutate(
    child_abroad = dplyr::if_else(
      rowSums(dplyr::across(dplyr::starts_with("P1HH39_"),
                            ~ as.numeric(.x == 6)),
              na.rm = TRUE) > 0, 1L, 0L),

    # CORRECT: use P1HH42_* for child-specific OFW status
    child_ofw = dplyr::if_else(
      rowSums(dplyr::across(dplyr::starts_with("P1HH42_"),
                            ~ as.numeric(.x == 1)),
              na.rm = TRUE) > 0, 1L, 0L),

    transnational_hh = dplyr::if_else(
      child_abroad == 1L | child_ofw == 1L, 1L, 0L)
  )

cat("Wave 1 transnational_hh:\n")
table(lsahp_w1$transnational_hh, useNA = "ifany")
```

## Mediators

```{r}
#| label: w1-mediators

# Contact frequency: minimum J10 code across children (lower = more frequent)
lsahp_w1 <- lsahp_w1 |>
  dplyr::mutate(
    contact_child_freq = {
      j10_mat <- dplyr::across(
        dplyr::starts_with("P1SJ10_"),
        ~ dplyr::if_else(.x %in% 0:4, as.numeric(.x), NA_real_)
      ) |> as.matrix()
      apply(j10_mat, 1, function(x) {
        if (all(is.na(x))) NA_real_ else min(x, na.rm = TRUE)
      })
    }
  )

# Contact satisfaction: binary (1 = satisfied, 0 = not)
# CORRECT coding: {1,2} = satisfied, {3} = not satisfied
lsahp_w1 <- lsahp_w1 |>
  dplyr::mutate(
    contact_child_satisfied = dplyr::case_when(
      P1SJ26 %in% c(1, 2) ~ 1L,
      P1SJ26 == 3          ~ 0L,
      TRUE                 ~ NA_integer_
    )
  )
```

## Distance to Farthest Migrant Child

```{r}
#| label: w1-distance
#| cache: true

library(rnaturalearth); library(sf); library(geosphere)
sf::sf_use_s2(FALSE)

ref_lon <- 120.9842; ref_lat <- 14.5995

world <- rnaturalearth::ne_countries(scale = "medium", returnclass = "sf")
world_centroids <- world |>
  dplyr::mutate(centroid = sf::st_centroid(geometry)) |>
  dplyr::mutate(lon = sf::st_coordinates(centroid)[,"X"],
                lat = sf::st_coordinates(centroid)[,"Y"]) |>
  dplyr::select(iso_a2, name, lon, lat)

country_vars_w1 <- lsahp_w1 |>
  dplyr::select(dplyr::starts_with("P1HH43_")) |> names()

country_codes_w1 <- lsahp_w1 |>
  dplyr::select(dplyr::all_of(country_vars_w1)) |>
  tidyr::pivot_longer(cols = dplyr::everything(), values_to = "country_code") |>
  dplyr::filter(!is.na(country_code), country_code != "") |>
  dplyr::distinct(country_code) |>
  dplyr::left_join(world_centroids, by = c("country_code" = "iso_a2"))

lsahp_w1 <- lsahp_w1 |>
  dplyr::rowwise() |>
  dplyr::mutate(
    max_child_distance_km = {
      codes <- c_across(dplyr::all_of(country_vars_w1))
      codes <- codes[!is.na(codes) & codes != ""]
      if (length(codes) == 0) { NA_real_ } else {
        coords <- country_codes_w1 |>
          dplyr::filter(country_code %in% codes, !is.na(lon), !is.na(lat))
        if (nrow(coords) == 0) { NA_real_ } else {
          max(geosphere::distHaversine(
            matrix(c(ref_lon, ref_lat), ncol = 2),
            as.matrix(coords[, c("lon","lat")])
          )) / 1000
        }
      }
    }
  ) |>
  dplyr::ungroup()
```

## Controls and IPW Covariates

```{r}
#| label: w1-controls

lsahp_w1 <- lsahp_w1 |>
  dplyr::mutate(
    urban      = dplyr::case_when(P1MX_URBAN == 1 ~ 1L,
                                  P1MX_URBAN == 2 ~ 0L, TRUE ~ NA_integer_),
    age        = as.numeric(P1SA2),
    sex        = dplyr::case_when(P1SA3 == 1 ~ 0L, P1SA3 == 2 ~ 1L,
                                  TRUE ~ NA_integer_),
    educhsplus = dplyr::case_when(P1SQ16_EDUC %in% c(2,3) ~ 1L,
                                  P1SQ16_EDUC == 1 ~ 0L, TRUE ~ NA_integer_),
    wealthq    = as.integer(P1MX_WINDEX),
    poor       = dplyr::case_when(P1MX_WINDEX == 1 ~ 1L,
                                  P1MX_WINDEX %in% 2:5 ~ 0L, TRUE ~ NA_integer_),
    health_3cat = dplyr::case_when(P1SB1 %in% c(1,2) ~ 0L, P1SB1 == 3 ~ 1L,
                                   P1SB1 %in% c(4,5) ~ 2L, TRUE ~ NA_integer_),
    marital    = dplyr::case_when(P1SA27 == 2 ~ 1L,
                                  P1SA27 %in% c(1,3,4,5) ~ 0L, TRUE ~ NA_integer_),
    n_children = as.numeric(P1HH30),
    remittance = dplyr::case_when(P1SF2_08 == 1 ~ 1L, P1SF2_08 == 2 ~ 0L,
                                  TRUE ~ NA_integer_),
    internet   = dplyr::case_when(P1SH16 == 1 ~ 1L, P1SH16 == 2 ~ 0L,
                                  TRUE ~ NA_integer_),
    cellphone  = dplyr::case_when(P1SH20 == 1 ~ 1L, P1SH20 == 2 ~ 0L,
                                  TRUE ~ NA_integer_),
    cognitive  = as.numeric(P1SL_TOTAL),
    social_org = dplyr::case_when(P1SH4 == 1 | P1SH6 == 1 ~ 1L, TRUE ~ 0L)
  )
```

------------------------------------------------------------------------

# Wave 2 Variables

```{r}
#| label: w2-all

lsahp_w2 <- lsahp2 |>
  dplyr::mutate(id = P1ID1)

# ── Binary loneliness ────────────────────────────────────────────────────────
lsahp_w2 <- lsahp_w2 |>
  dplyr::mutate(
    lonely_w2 = dplyr::case_when(
      P2SD1_06 == 3          ~ 1L,
      P2SD1_06 %in% c(1, 2) ~ 0L,
      TRUE                   ~ NA_integer_
    )
  )

# ── 3-item scale ─────────────────────────────────────────────────────────────
# Wave 2 uses same 0-4 scale as Wave 1 (confirmed from codebook)
# NA = 2,567 at Wave 2 vs 775 at Wave 1 — attrition driven, not scale change
lsahp_w2 <- lsahp_w2 |>
  dplyr::mutate(
    lonely_3item_w2 = dplyr::if_else(
      rowSums(dplyr::across(c(P2SH9_1, P2SH9_2, P2SH9_3),
                            ~ is.na(.x))) == 0,
      rowMeans(dplyr::across(c(P2SH9_1, P2SH9_2, P2SH9_3))),
      NA_real_
    )
  )

# ── Transnational exposure ───────────────────────────────────────────────────
lsahp_w2 <- lsahp_w2 |>
  dplyr::mutate(
    child_abroad_w2 = dplyr::if_else(
      rowSums(dplyr::across(dplyr::starts_with("P2HH39_"),
                            ~ as.numeric(.x == 6)), na.rm = TRUE) > 0, 1L, 0L),

    # CORRECTED: P2HH42_* not P2HH14_* for child-specific OFW
    child_ofw_w2 = dplyr::if_else(
      rowSums(dplyr::across(dplyr::starts_with("P2HH42_"),
                            ~ as.numeric(.x == 1)), na.rm = TRUE) > 0, 1L, 0L),

    transnational_hh_w2 = dplyr::if_else(
      child_abroad_w2 == 1L | child_ofw_w2 == 1L, 1L, 0L)
  )

# ── Contact frequency ────────────────────────────────────────────────────────
lsahp_w2 <- lsahp_w2 |>
  dplyr::mutate(
    contact_child_freq_w2 = {
      j10_mat <- dplyr::across(
        dplyr::starts_with("P2SJ10"),
        ~ dplyr::if_else(.x %in% 0:4, as.numeric(.x), NA_real_)
      ) |> as.matrix()
      apply(j10_mat, 1, function(x) {
        if (all(is.na(x))) NA_real_ else min(x, na.rm = TRUE)
      })
    }
  )

# ── Contact satisfaction ─────────────────────────────────────────────────────
# CORRECTED: {1,2} = satisfied, {3} = not satisfied (same as Wave 1)
lsahp_w2 <- lsahp_w2 |>
  dplyr::mutate(
    contact_child_satisfied_w2 = dplyr::case_when(
      P2SJ26 %in% c(1, 2) ~ 1L,
      P2SJ26 == 3          ~ 0L,
      TRUE                 ~ NA_integer_
    )
  )

# ── Controls ─────────────────────────────────────────────────────────────────
lsahp_w2 <- lsahp_w2 |>
  dplyr::mutate(
    urban_w2 = dplyr::case_when(P2MX_URBAN == 1 ~ 1L,
                                P2MX_URBAN == 0 ~ 0L, TRUE ~ NA_integer_),
    age_w2   = as.numeric(P2SA2),
    sex_w2   = dplyr::case_when(P2SA3 == 1 ~ 0L, P2SA3 == 2 ~ 1L,
                                TRUE ~ NA_integer_),
    educhsplus_w2 = dplyr::case_when(w2op_educ3 %in% c(2,3) ~ 1L,
                                     w2op_educ3 == 1 ~ 0L, TRUE ~ NA_integer_),
    wealthq_w2 = as.integer(P2MX_WINDEX_A),
    poor_w2    = dplyr::case_when(P2MX_WINDEX_A == 1 ~ 1L,
                                  P2MX_WINDEX_A %in% 2:5 ~ 0L, TRUE ~ NA_integer_),
    health_3cat_w2 = dplyr::case_when(P2SB1 %in% c(1,2) ~ 0L,
                                      P2SB1 == 3 ~ 1L,
                                      P2SB1 %in% c(4,5) ~ 2L, TRUE ~ NA_integer_)
  )
```

## Wave 2 Distance

```{r}
#| label: w2-distance
#| cache: true

sf::sf_use_s2(FALSE)

world_centroids_n3 <- world |>
  dplyr::mutate(centroid = sf::st_centroid(geometry)) |>
  dplyr::mutate(lon = sf::st_coordinates(centroid)[,"X"],
                lat = sf::st_coordinates(centroid)[,"Y"]) |>
  dplyr::select(iso_n3, name, lon, lat) |>
  dplyr::mutate(iso_n3 = as.numeric(iso_n3))

country_vars_w2 <- lsahp_w2 |>
  dplyr::select(dplyr::starts_with("P2HH43_")) |> names()

country_codes_w2 <- lsahp_w2 |>
  dplyr::select(dplyr::all_of(country_vars_w2)) |>
  tidyr::pivot_longer(cols = dplyr::everything(), values_to = "country_code") |>
  dplyr::mutate(country_code = as.numeric(country_code)) |>
  dplyr::filter(!is.na(country_code), country_code > 0) |>
  dplyr::distinct(country_code) |>
  dplyr::left_join(world_centroids_n3, by = c("country_code" = "iso_n3"))

lsahp_w2 <- lsahp_w2 |>
  dplyr::rowwise() |>
  dplyr::mutate(
    max_child_distance_km_w2 = {
      codes <- as.numeric(c_across(dplyr::all_of(country_vars_w2)))
      codes <- codes[!is.na(codes) & codes > 0]
      if (length(codes) == 0) { NA_real_ } else {
        coords <- country_codes_w2 |>
          dplyr::filter(country_code %in% codes, !is.na(lon), !is.na(lat))
        if (nrow(coords) == 0) { NA_real_ } else {
          max(geosphere::distHaversine(
            matrix(c(ref_lon, ref_lat), ncol = 2),
            as.matrix(coords[, c("lon","lat")])
          )) / 1000
        }
      }
    }
  ) |>
  dplyr::ungroup()
```

------------------------------------------------------------------------

# Panel Construction

```{r}
#| label: panel-build

panel_ids <- intersect(lsahp_w1$id, lsahp_w2$id)
cat("Balanced panel n =", length(panel_ids), "\n")

lsahp_w1p <- lsahp_w1 |> dplyr::filter(id %in% panel_ids)
lsahp_w2p <- lsahp_w2 |> dplyr::filter(id %in% panel_ids)

# Long format
lsahp_w1_long <- lsahp_w1p |>
  dplyr::transmute(
    id, wave = 1L,
    lonely = as.numeric(lonely_w1), lonely_3item,
    transnational_hh, contact_child_freq, contact_child_satisfied,
    max_child_distance_km, urban, age, sex, educhsplus,
    wealthq, poor, health_3cat,
    strata = P1MX_STRATA, psu = P1MX_PSU, wgt = P1MX_WGTU
  )

lsahp_w2_long <- lsahp_w2p |>
  dplyr::transmute(
    id, wave = 2L,
    lonely = as.numeric(lonely_w2), lonely_3item = lonely_3item_w2,
    transnational_hh = transnational_hh_w2,
    contact_child_freq = contact_child_freq_w2,
    contact_child_satisfied = contact_child_satisfied_w2,
    max_child_distance_km = max_child_distance_km_w2,
    urban = urban_w2, age = age_w2, sex = sex_w2,
    educhsplus = educhsplus_w2, wealthq = wealthq_w2,
    poor = poor_w2, health_3cat = health_3cat_w2,
    strata = P1MX_STRATA, psu = P1MX_PSU, wgt = P2MX_WGT1
  )

lsahp_long <- dplyr::bind_rows(lsahp_w1_long, lsahp_w2_long)

# Wide format
lsahp_wide <- lsahp_long |>
  dplyr::select(id, wave, lonely, lonely_3item, transnational_hh,
                contact_child_freq, contact_child_satisfied,
                max_child_distance_km) |>
  tidyr::pivot_wider(names_from = wave,
                     values_from = c(lonely, lonely_3item, transnational_hh,
                                     contact_child_freq, contact_child_satisfied,
                                     max_child_distance_km),
                     names_sep = "_w")

# Controls wide
controls_wide <- lsahp_long |>
  dplyr::select(id, wave, poor, urban, health_3cat) |>
  tidyr::pivot_wider(names_from = wave,
                     values_from = c(poor, urban, health_3cat),
                     names_sep = "_w")

lsahp_wide <- lsahp_wide |>
  dplyr::left_join(controls_wide, by = "id")

# Transition indicators and first differences
lsahp_wide <- lsahp_wide |>
  dplyr::mutate(
    trans_transition = dplyr::case_when(
      transnational_hh_w1 == 0 & transnational_hh_w2 == 0 ~ "Stayed non-transnational",
      transnational_hh_w1 == 0 & transnational_hh_w2 == 1 ~ "Entered transnational",
      transnational_hh_w1 == 1 & transnational_hh_w2 == 1 ~ "Stayed transnational",
      transnational_hh_w1 == 1 & transnational_hh_w2 == 0 ~ "Exited transnational",
      TRUE ~ NA_character_
    ),
    entered = dplyr::if_else(
      transnational_hh_w1 == 0 & transnational_hh_w2 == 1, 1L, 0L),
    exited  = dplyr::if_else(
      transnational_hh_w1 == 1 & transnational_hh_w2 == 0, 1L, 0L),
    # First differences
    d_lonely        = as.integer(lonely_w2) - as.integer(lonely_w1),
    d_lonely3       = lonely_3item_w2 - lonely_3item_w1,
    d_contact_sat   = contact_child_satisfied_w2 - contact_child_satisfied_w1,
    d_contact_freq  = contact_child_freq_w2 - contact_child_freq_w1,
    d_dist_km       = max_child_distance_km_w2 - max_child_distance_km_w1,
    d_dist_1000km   = d_dist_km / 1000,
    d_poor          = poor_w2   - poor_w1,
    d_urban         = urban_w2  - urban_w1,
    d_health        = health_3cat_w2 - health_3cat_w1,
    # Interaction term (satisfaction × distance)
    d_sat_x_dist    = d_contact_sat * d_dist_1000km,
    # Became lonely
    became_lonely   = dplyr::if_else(d_lonely == 1, 1L, 0L)
  )

cat("Wide n =", nrow(lsahp_wide), "\n")
cat("\nTransition distribution:\n")
table(lsahp_wide$trans_transition, useNA = "ifany")
```

------------------------------------------------------------------------

# Missingness Audit

```{r}
#| label: missing-audit

lsahp_wide |>
  dplyr::summarise(
    n_total                = dplyr::n(),
    miss_d_lonely          = sum(is.na(d_lonely)),
    miss_d_lonely3         = sum(is.na(d_lonely3)),
    miss_d_contact_sat     = sum(is.na(d_contact_sat)),
    miss_d_dist_1000km     = sum(is.na(d_dist_1000km)),
    miss_d_poor            = sum(is.na(d_poor)),
    miss_d_urban           = sum(is.na(d_urban)),
    miss_d_health          = sum(is.na(d_health)),
    n_tier1_cont           = sum(!is.na(d_lonely3) & !is.na(entered) &
                                 !is.na(d_poor) & !is.na(d_urban) &
                                 !is.na(d_health)),
    n_tier2_contact        = sum(!is.na(d_lonely3) & !is.na(entered) &
                                 !is.na(d_contact_sat) &
                                 !is.na(d_poor) & !is.na(d_urban) &
                                 !is.na(d_health)),
    n_tier3_distance       = sum(!is.na(d_lonely3) & !is.na(entered) &
                                 !is.na(d_dist_1000km) &
                                 !is.na(d_poor) & !is.na(d_urban) &
                                 !is.na(d_health)),
    n_tier3_interaction    = sum(!is.na(d_lonely3) & !is.na(entered) &
                                 !is.na(d_contact_sat) & !is.na(d_dist_1000km) &
                                 !is.na(d_poor) & !is.na(d_urban) &
                                 !is.na(d_health))
  ) |>
  tidyr::pivot_longer(dplyr::everything()) |>
  knitr::kable(caption = "Missingness audit and analytic sample sizes by tier")

# CRITICAL: confirm zero enterers in distance subsample
cat("\n=== Enterer breakdown in distance subsample ===\n")
lsahp_wide |>
  dplyr::filter(!is.na(d_lonely3) & !is.na(d_dist_1000km) &
                !is.na(d_poor) & !is.na(d_urban) & !is.na(d_health)) |>
  dplyr::count(entered, exited) |>
  dplyr::mutate(group = dplyr::case_when(
    entered == 1 ~ "Enterers (0→1)",
    exited  == 1 ~ "Exiters  (1→0)",
    TRUE         ~ "Stayers (no change)"
  )) |>
  knitr::kable(caption = "Transition groups in distance subsample (n=395). NOTE: zero enterers — distance analysis is within already-transnational households only.")
```

------------------------------------------------------------------------

# Descriptives

## Survey Designs

```{r}
#| label: survey-designs

library(survey)

des_w1 <- svydesign(ids = ~psu, strata = ~strata, weights = ~wgt,
                    data = dplyr::filter(lsahp_long, wave == 1, !is.na(wgt)),
                    nest = TRUE)

des_w2 <- svydesign(ids = ~psu, strata = ~strata, weights = ~wgt,
                    data = dplyr::filter(lsahp_long, wave == 2, !is.na(wgt)),
                    nest = TRUE)

design_w1_vars <- lsahp_long |>
  dplyr::filter(wave == 1) |>
  dplyr::select(id, strata, psu, wgt) |>
  dplyr::distinct()

lsahp_wide_des <- lsahp_wide |>
  dplyr::left_join(design_w1_vars, by = "id")

des_wide <- svydesign(ids = ~psu, strata = ~strata, weights = ~wgt,
                      data = lsahp_wide_des, nest = TRUE)

svymean_ci <- function(var, design, level = 0.95) {
  m  <- survey::svymean(var, design, na.rm = TRUE)
  ci <- suppressMessages(confint(m, level = level))
  data.frame(mean = as.numeric(coef(m)),
             ci_l = as.numeric(ci[,1]),
             ci_u = as.numeric(ci[,2]))
}
```

## Figure 1: Loneliness Prevalence

```{r}
#| label: fig-lonely
#| fig-cap: "Weighted prevalence of frequent loneliness (often = 1 on CES-D) by survey wave."

prev_lon <- dplyr::bind_rows(
  svymean_ci(~lonely, des_w1) |> dplyr::mutate(Wave = "Wave 1"),
  svymean_ci(~lonely, des_w2) |> dplyr::mutate(Wave = "Wave 2")
) |> dplyr::mutate(label = scales::percent(mean, accuracy = 0.1))

p_lon <- ggplot(prev_lon, aes(x = Wave, y = mean)) +
  geom_col(fill = LMUGreen, color = "#006B2C", width = 0.6) +
  geom_errorbar(aes(ymin = ci_l, ymax = ci_u), width = 0.12, linewidth = 0.9) +
  geom_text(aes(y = ci_u + 0.015, label = label), size = 6, fontface = "bold") +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1),
                     limits = c(0, 0.20)) +
  labs(title = "Prevalence of Frequent Loneliness, by Wave (Weighted)",
       x = "Survey wave", y = "Proportion reporting frequent loneliness") +
  theme_lmu()

save_plot(p_lon, "01_lonely_prevalence")
p_lon
```

## Figure 2: Transnational Prevalence

```{r}
#| label: fig-transnational
#| fig-cap: "Weighted prevalence of transnational households by survey wave."

prev_tn <- dplyr::bind_rows(
  svymean_ci(~transnational_hh, des_w1) |> dplyr::mutate(Wave = "Wave 1"),
  svymean_ci(~transnational_hh, des_w2) |> dplyr::mutate(Wave = "Wave 2")
) |> dplyr::mutate(label = scales::percent(mean, accuracy = 0.1))

p_tn <- ggplot(prev_tn, aes(x = Wave, y = mean)) +
  geom_col(fill = LMUGreen, color = "#006B2C", width = 0.6) +
  geom_errorbar(aes(ymin = ci_l, ymax = ci_u), width = 0.12, linewidth = 0.9) +
  geom_text(aes(y = ci_u + 0.03, label = label), size = 6, fontface = "bold") +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1),
                     limits = c(0, 0.30)) +
  labs(title = "Prevalence of Transnational Households, by Wave (Weighted)",
       x = "Survey wave", y = "Proportion in a transnational household") +
  theme_lmu()

save_plot(p_tn, "02_transnational_prevalence")
p_tn
```

## Figure 3: Transition Distribution

```{r}
#| label: fig-transitions
#| fig-cap: "Weighted distribution of household transitions between Wave 1 and Wave 2."

m_trans <- svymean(~factor(trans_transition), des_wide, na.rm = TRUE)
ci_trans <- suppressMessages(confint(m_trans))

trans_df <- data.frame(
  Transition = gsub("^factor\\(trans_transition\\)", "", names(coef(m_trans))),
  mean = as.numeric(coef(m_trans)),
  ci_l = as.numeric(ci_trans[,1]),
  ci_u = as.numeric(ci_trans[,2])
) |> dplyr::mutate(label = scales::percent(mean, accuracy = 0.1))

p_trans <- ggplot(trans_df, aes(x = reorder(Transition, -mean), y = mean)) +
  geom_col(fill = LMUGreen, color = "#006B2C", width = 0.7) +
  geom_errorbar(aes(ymin = ci_l, ymax = ci_u), width = 0.12, linewidth = 0.9) +
  geom_text(aes(y = ci_u + 0.03, label = label), size = 6, fontface = "bold") +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1),
                     limits = c(0, 0.90)) +
  labs(title = "Transnational Household Transitions (Weighted)",
       x = "Transition group (Wave 1 → Wave 2)",
       y = "Proportion of the sample") +
  theme_lmu() +
  theme(axis.text.x = element_text(angle = 15, hjust = 1))

save_plot(p_trans, "03_transitions", w = 10, h = 6)
p_trans
```

## Figure 4: Becoming Lonely by Transition

```{r}
#| label: fig-became-lonely
#| fig-cap: "Weighted proportion becoming frequently lonely by transition group."

svy_mean_by_group <- function(design, yvar, gvar) {
  levs <- levels(factor(design$variables[[gvar]]))
  dplyr::bind_rows(lapply(levs, function(L) {
    dsub <- subset(design, design$variables[[gvar]] == L)
    m    <- svymean(as.formula(paste0("~", yvar)), dsub, na.rm = TRUE)
    ci   <- suppressMessages(confint(m))
    data.frame(group = L, mean = as.numeric(coef(m)),
               ci_l = as.numeric(ci[,1]), ci_u = as.numeric(ci[,2]))
  }))
}

bl_df <- svy_mean_by_group(des_wide, "became_lonely", "trans_transition") |>
  dplyr::rename(Transition = group) |>
  dplyr::mutate(label = scales::percent(mean, accuracy = 0.1))

p_bl <- ggplot(bl_df, aes(x = reorder(Transition, -mean), y = mean)) +
  geom_col(fill = LMUGreen, color = "#006B2C", width = 0.7) +
  geom_errorbar(aes(ymin = ci_l, ymax = ci_u), width = 0.12, linewidth = 0.9) +
  geom_text(aes(y = ci_u + 0.04, label = label), size = 6, fontface = "bold") +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1),
                     limits = c(0, 0.45)) +
  labs(title = "Percent Becoming Lonely by Transnational Transition (Weighted)",
       x = "Transition group (Wave 1 → Wave 2)",
       y = "Proportion who became frequently lonely") +
  theme_lmu() +
  theme(axis.text.x = element_text(angle = 15, hjust = 1))

save_plot(p_bl, "04_became_lonely", w = 10, h = 6)
p_bl
```

------------------------------------------------------------------------

# TIER 1: Causal Analysis (H1)

## IPW — Propensity Score Estimation

```{r}
#| label: ipw-setup

library(WeightIt); library(cobalt)

baseline_vars <- lsahp_w1 |>
  dplyr::transmute(
    id,
    age_w1        = age,
    sex_w1        = sex,
    educ_w1       = educhsplus,
    wealth_w1     = wealthq,
    marital_w1    = marital,
    health_w1     = health_3cat,
    n_children_w1 = n_children,
    # remittance_w1 REMOVED — 97% missing, not usable
    internet_w1   = internet,
    cellphone_w1  = cellphone,
    cognitive_w1  = cognitive,
    lonely3_w1    = lonely_3item,
    social_org_w1 = social_org
    # urban_w1 and poor_w1 already in lsahp_wide_des — not joined here
  )

lsahp_wide_des2 <- lsahp_wide_des |>
  dplyr::left_join(baseline_vars, by = "id")

# IPW sample: baseline non-transnational only
lsahp_ipw <- lsahp_wide_des2 |>
  dplyr::filter(transnational_hh_w1 == 0) |>
  dplyr::mutate(
    d_dist_1000km = d_dist_km / 1000,
    d_sat_x_dist  = d_contact_sat * d_dist_1000km
  )

# Mean imputation for PS covariates with <20% missingness
# Standard practice for propensity score estimation
# Reported in methods: "Missing values on PS covariates imputed
# to the covariate mean prior to propensity score estimation"
lsahp_ipw <- lsahp_ipw |>
  dplyr::mutate(
    educ_w1       = dplyr::if_else(
      is.na(educ_w1),
      mean(educ_w1, na.rm = TRUE), as.numeric(educ_w1)),
    health_w1     = dplyr::if_else(
      is.na(health_w1),
      mean(health_w1, na.rm = TRUE), as.numeric(health_w1)),
    n_children_w1 = dplyr::if_else(
      is.na(n_children_w1),
      mean(n_children_w1, na.rm = TRUE), as.numeric(n_children_w1)),
    cognitive_w1  = dplyr::if_else(
      is.na(cognitive_w1),
      mean(cognitive_w1, na.rm = TRUE), as.numeric(cognitive_w1)),
    lonely3_w1    = dplyr::if_else(
      is.na(lonely3_w1),
      mean(lonely3_w1, na.rm = TRUE), as.numeric(lonely3_w1))
  )

cat("IPW sample: n =", nrow(lsahp_ipw),
    "| Enterers:", sum(lsahp_ipw$entered, na.rm = TRUE),
    "| Stayers:",  sum(lsahp_ipw$entered == 0, na.rm = TRUE), "\n")

# Confirm no remaining missingness on PS covariates
ps_vars <- c("age_w1","sex_w1","educ_w1","wealth_w1","urban_w1",
             "marital_w1","health_w1","n_children_w1",
             "internet_w1","cellphone_w1",
             "cognitive_w1","lonely3_w1","social_org_w1",
             "poor_w1")

remaining_miss <- lsahp_ipw |>
  dplyr::select(dplyr::all_of(ps_vars)) |>
  dplyr::summarise(dplyr::across(dplyr::everything(), ~ sum(is.na(.x)))) |>
  tidyr::pivot_longer(dplyr::everything()) |>
  dplyr::filter(value > 0)

if (nrow(remaining_miss) == 0) {
  cat("No missing values remaining in PS covariates.\n")
} else {
  print(remaining_miss)
}
```

```{r}
#| label: ipw-estimate

# Updated formula: remittance_w1 removed (97% missing)
ps_formula <- entered ~
  age_w1 + sex_w1 + educ_w1 + wealth_w1 + urban_w1 +
  marital_w1 + health_w1 + n_children_w1 +
  poor_w1 + internet_w1 + cellphone_w1 +
  cognitive_w1 + lonely3_w1 + social_org_w1

w_out <- WeightIt::weightit(
  ps_formula,
  data     = lsahp_ipw,
  method   = "ps",
  estimand = "ATT"
)

summary(w_out)

# Combine IPW with survey weights
lsahp_ipw <- lsahp_ipw |>
  dplyr::mutate(
    ipw          = w_out$weights,
    # Check if trimming needed
    p99          = quantile(w_out$weights, 0.99, na.rm = TRUE),
    ipw_trimmed  = pmin(ipw, p99),
    wgt_combined = ipw_trimmed * wgt
  )

cat("\nIPW weight summary:\n")
summary(lsahp_ipw$ipw)
cat("\n99th percentile:", round(quantile(lsahp_ipw$ipw, 0.99), 3),
    "— trimming", 
    ifelse(quantile(lsahp_ipw$ipw, 0.99) > 5, "RECOMMENDED", "not needed"), "\n")
```

## Balance Assessment

```{r}
#| label: fig-balance
#| fig-cap: "Covariate balance before and after IPW. All adjusted SMDs should fall within the dashed lines at ±0.10."

bal_tab <- cobalt::bal.tab(
  w_out,
  stats      = c("mean.diffs", "variance.ratios"),
  thresholds = c(m = 0.1)
)
print(bal_tab)

cobalt::love.plot(
  w_out,
  threshold = 0.1,
  colors    = c("#C0392B", LMUGreen),
  title     = "Covariate Balance Before and After IPW"
)
```

## Trim and Combine Weights

```{r}
#| label: ipw-trim

p99 <- quantile(w_out$weights, 0.99, na.rm = TRUE)

lsahp_ipw <- lsahp_ipw |>
  dplyr::mutate(
    ipw          = w_out$weights,
    ipw_trimmed  = pmin(ipw, p99),
    wgt_combined = ipw_trimmed * wgt
  )

cat("Weight summary after trimming at 99th percentile:\n")
summary(lsahp_ipw$ipw_trimmed)
# Correct ESS reporting
ess_treated  <- sum(lsahp_ipw$ipw[lsahp_ipw$entered == 1])^2 /
                sum(lsahp_ipw$ipw[lsahp_ipw$entered == 1]^2)

ess_controls <- sum(lsahp_ipw$ipw[lsahp_ipw$entered == 0])^2 /
                sum(lsahp_ipw$ipw[lsahp_ipw$entered == 0]^2)

cat("Effective sample size:\n")
cat("  Treated (enterers):", round(ess_treated, 0), "\n")
cat("  Controls (stayers):", round(ess_controls, 0), "\n")
cat("  Total:             ", round(ess_treated + ess_controls, 0), "\n")
cat("\nControl weight range:",
    round(min(lsahp_ipw$ipw[lsahp_ipw$entered == 0]), 3), "to",
    round(max(lsahp_ipw$ipw[lsahp_ipw$entered == 0]), 3), "\n")
cat("Trimming required: NO (max weight = 0.440)\n")
```

## Tier 1 Models

```{r}
#| label: tier1-models

# Plain FD (comparison baseline) — IPW sample only
des_plain <- svydesign(ids = ~psu, strata = ~strata, weights = ~wgt,
                       data = lsahp_ipw, nest = TRUE)

# IPW-weighted FD (main causal estimate)
des_ipw <- svydesign(ids = ~psu, strata = ~strata, weights = ~wgt_combined,
                     data = lsahp_ipw, nest = TRUE)

# PRIMARY OUTCOME: 3-item continuous scale
# NOTE: exited removed — IPW sample is baseline non-transnational only
#       so exited = 0 for everyone (no variation, dropped automatically)
m1_plain_cont <- svyglm(
  d_lonely3 ~ entered + d_poor + d_urban + d_health,
  design = des_plain, na.action = na.omit)

m1_ipw_cont <- svyglm(
  d_lonely3 ~ entered + d_poor + d_urban + d_health,
  design = des_ipw, na.action = na.omit)

# ROBUSTNESS: binary outcome
m1_plain_bin <- svyglm(
  d_lonely ~ entered + d_poor + d_urban + d_health,
  design = des_plain, na.action = na.omit)

m1_ipw_bin <- svyglm(
  d_lonely ~ entered + d_poor + d_urban + d_health,
  design = des_ipw, na.action = na.omit)

cat("=== TIER 1: Main causal model (3-item, IPW) ===\n")
summary(m1_ipw_cont)

cat("\n=== TIER 1: Robustness (binary, IPW) ===\n")
summary(m1_ipw_bin)

cat("\n=== Plain FD vs IPW comparison ===\n")
data.frame(
  Model      = c("Plain FD", "FD + IPW"),
  Coef_3item = c(coef(m1_plain_cont)["entered"],
                 coef(m1_ipw_cont)["entered"]),
  SE_3item   = c(sqrt(vcov(m1_plain_cont)["entered","entered"]),
                 sqrt(vcov(m1_ipw_cont)["entered","entered"]))
) |>
  dplyr::mutate(
    ci_lo = Coef_3item - 1.96 * SE_3item,
    ci_hi = Coef_3item + 1.96 * SE_3item,
    across(where(is.numeric), ~ round(.x, 3))
  ) |>
  print()

cat("\nTier 1 sample sizes:\n")
cat("Plain FD n =", nrow(model.frame(m1_plain_cont)), "\n")
cat("IPW n      =", nrow(model.frame(m1_ipw_cont)), "\n")

# Force print the comparison table
comparison <- data.frame(
  Model      = c("Plain FD", "FD + IPW"),
  Coef_3item = c(coef(m1_plain_cont)["entered"],
                 coef(m1_ipw_cont)["entered"]),
  SE_3item   = c(sqrt(vcov(m1_plain_cont)["entered","entered"]),
                 sqrt(vcov(m1_ipw_cont)["entered","entered"]))
) |>
  dplyr::mutate(
    ci_lo = round(Coef_3item - 1.96 * SE_3item, 3),
    ci_hi = round(Coef_3item + 1.96 * SE_3item, 3),
    Coef_3item = round(Coef_3item, 3),
    SE_3item   = round(SE_3item, 3)
  )

print(comparison)

# Diagnose why n=2628 not ~3291
# The IPW sample has 4795 rows — what is causing 2167 to drop?
cat("\nMissingness in Tier 1 model variables (IPW sample):\n")
lsahp_ipw |>
  dplyr::summarise(
    n_total      = dplyr::n(),
    miss_d_lonely3 = sum(is.na(d_lonely3)),
    miss_entered   = sum(is.na(entered)),
    miss_d_poor    = sum(is.na(d_poor)),
    miss_d_urban   = sum(is.na(d_urban)),
    miss_d_health  = sum(is.na(d_health)),
    n_complete     = sum(
      !is.na(d_lonely3) & !is.na(entered) &
      !is.na(d_poor) & !is.na(d_urban) & !is.na(d_health)
    )
  ) |>
  tidyr::pivot_longer(dplyr::everything()) |>
  print()
```

## Tier 1 Robustness Comparison

```{r}
#| label: tier1-compare

data.frame(
  Specification = c(
    "FD plain — 3-item",
    "FD + IPW — 3-item (PRIMARY)",
    "FD plain — binary",
    "FD + IPW — binary (ROBUSTNESS)"
  ),
  Entered = c(
    coef(m1_plain_cont)["entered"],
    coef(m1_ipw_cont)["entered"],
    coef(m1_plain_bin)["entered"] * 100,
    coef(m1_ipw_bin)["entered"] * 100
  ),
  Units = c(
    "scale points", "scale points", "pct pts", "pct pts"
  )
) |> knitr::kable(digits = 3,
                  caption = "Robustness check: plain FD vs IPW-weighted FD. Similar estimates indicate selection on observables does not drive the main finding.")
```

------------------------------------------------------------------------

# TIER 2: Associational / Mechanism Analysis (H2)

```{r}
#| label: tier2-models

# NOTE: n = 3,075 — respondents with non-resident children and valid contact data
# These are within-person associations; reverse causality cannot be ruled out

# Full sample survey design for Tier 2 (not IPW-restricted)
lsahp_wide_des3 <- lsahp_wide_des |>
  dplyr::mutate(d_dist_1000km = d_dist_km / 1000,
                d_sat_x_dist  = d_contact_sat * d_dist_1000km)

des_full <- svydesign(ids = ~psu, strata = ~strata, weights = ~wgt,
                      data = lsahp_wide_des3, nest = TRUE)

# M2a: + Contact frequency (mechanism: does frequency explain entry effect?)
m2a <- svyglm(
  d_lonely3 ~ entered + exited +
    d_contact_freq +
    d_poor + d_urban + d_health,
  design = des_full, na.action = na.omit)

# M2b: + Contact satisfaction (mechanism: does quality explain entry effect?)
m2b <- svyglm(
  d_lonely3 ~ entered + exited +
    d_contact_sat +
    d_poor + d_urban + d_health,
  design = des_full, na.action = na.omit)

# M2c: Both contact variables
m2c <- svyglm(
  d_lonely3 ~ entered + exited +
    d_contact_sat + d_contact_freq +
    d_poor + d_urban + d_health,
  design = des_full, na.action = na.omit)

cat("=== TIER 2: Contact satisfaction mechanism ===\n")
summary(m2b)

cat("\n=== TIER 2: Both contact variables ===\n")
summary(m2c)
```

```{r}
# ── Tier 2 binary (to match original paper results) ─────────────────────────

m2b_bin <- svyglm(
  d_lonely ~ entered + exited +
    d_contact_sat +
    d_poor + d_urban + d_health,
  design = des_full, na.action = na.omit)

cat("=== TIER 2b BINARY: Contact satisfaction ===\n")
summary(m2b_bin)
cat("n =", nrow(model.frame(m2b_bin)),
    "| enterers =", sum(model.frame(m2b_bin)$entered), "\n")

# ── Tier 3: Already-transnational subsample ──────────────────────────────────

# Restrict to already-transnational at Wave 1 with distance recorded
lsahp_dist <- lsahp_wide_des3 |>
  dplyr::filter(
    transnational_hh_w1 == 1,
    !is.na(d_dist_1000km)
  )

cat("\nTier 3 subsample: n =", nrow(lsahp_dist), "\n")
cat("Transnational at W2:\n")
table(lsahp_dist$transnational_hh_w2)

des_dist <- svydesign(
  ids     = ~psu,
  strata  = ~strata,
  weights = ~wgt,
  data    = lsahp_dist,
  nest    = TRUE
)

# H3: Distance main effect
m3_dist <- svyglm(
  d_lonely3 ~ d_dist_1000km + d_poor + d_urban + d_health,
  design    = des_dist,
  na.action = na.omit
)

cat("\n=== TIER 3 (EXPLORATORY): Distance main effect ===\n")
summary(m3_dist)
cat("n =", nrow(model.frame(m3_dist)), "\n")

# H4: Satisfaction × Distance interaction
m4_interact <- svyglm(
  d_lonely3 ~ d_contact_sat * d_dist_1000km +
    d_poor + d_urban + d_health,
  design    = des_dist,
  na.action = na.omit
)

cat("\n=== TIER 3 (EXPLORATORY): Satisfaction × Distance ===\n")
summary(m4_interact)
cat("n =", nrow(model.frame(m4_interact)), "\n")

# Marginal effects: distance effect at each satisfaction level
library(marginaleffects)
mfx <- marginaleffects::avg_slopes(
  m4_interact,
  variables = "d_dist_1000km",
  by        = "d_contact_sat",
  newdata   = datagrid(d_contact_sat = c(-1, 0, 1))
)

cat("\n=== Marginal effect of distance by satisfaction level ===\n")
print(as.data.frame(mfx) |>
  dplyr::mutate(
    sat_label = dplyr::case_when(
      d_contact_sat == -1 ~ "Became less satisfied",
      d_contact_sat ==  0 ~ "No change",
      d_contact_sat ==  1 ~ "Became more satisfied"
    )
  ) |>
  dplyr::select(sat_label, estimate, conf.low, conf.high, p.value) |>
  dplyr::mutate(across(where(is.numeric), ~ round(.x, 4))))
```

# TIER 3: Exploratory Analysis (H3 + H4)

> **Important framing note.** The 395 respondents with recorded destination countries in both waves are **exclusively households already transnational at Wave 1 that remained transnational at Wave 2** — confirmed zero enterers in this subsample. The distance analysis does not identify the entry effect on loneliness. It answers a distinct question: *among already-transnational households, does a child moving farther away increase parental loneliness, and does contact satisfaction buffer this?* All Tier 3 findings are preliminary and exploratory.

## Distance Subsample

```{r}
#| label: tier3-subsample

# Restrict to already-transnational at Wave 1 with distance recorded
lsahp_dist <- lsahp_wide_des3 |>
  dplyr::filter(
    transnational_hh_w1 == 1,
    !is.na(d_dist_1000km)
  )

cat("Distance subsample: n =", nrow(lsahp_dist), "\n")
cat("All still transnational at W2:",
    all(lsahp_dist$transnational_hh_w2 == 1, na.rm = TRUE), "\n")
cat("Enterers (should be 0):",
    sum(lsahp_dist$entered == 1, na.rm = TRUE), "\n")

des_dist <- svydesign(
  ids     = ~psu,
  strata  = ~strata,
  weights = ~wgt,
  data    = lsahp_dist,
  nest    = TRUE
)
```

## H3: Distance Main Effect

```{r}
#| label: tier3-h3

# Among already-transnational households:
# does a child moving farther away increase loneliness?
m3_dist <- svyglm(
  d_lonely3 ~ d_dist_1000km + d_poor + d_urban + d_health,
  design    = des_dist,
  na.action = na.omit
)

cat("=== TIER 3 (EXPLORATORY): Distance main effect ===\n")
cat("n =", nrow(model.frame(m3_dist)),
    "| Already-transnational households only\n\n")
summary(m3_dist)
```

**Interpretation:** Distance alone is not significantly associated with loneliness change among already-transnational households (β = 0.009, p = 0.726). However, becoming poor is significantly associated with increased loneliness (β = 0.471, p = 0.033), suggesting economic vulnerability is a stronger driver of loneliness than geographic separation per se in this subsample.

## H4: Satisfaction × Distance Moderation

```{r}
#| label: tier3-h4

# EXPLORATORY: does satisfaction buffer the distance effect?
m4_interact <- svyglm(
  d_lonely3 ~ d_contact_sat * d_dist_1000km +
    d_poor + d_urban + d_health,
  design    = des_dist,
  na.action = na.omit
)

cat("=== TIER 3 (EXPLORATORY): Satisfaction × Distance ===\n")
cat("n =", nrow(model.frame(m4_interact)),
    "| Already-transnational only | Treat as preliminary\n\n")
summary(m4_interact)
```

```{r}
#| label: tier3-h4-table

# Key interaction coefficients
broom::tidy(m4_interact, conf.int = TRUE) |>
  dplyr::filter(term %in% c(
    "d_contact_sat",
    "d_dist_1000km",
    "d_contact_sat:d_dist_1000km"
  )) |>
  dplyr::mutate(
    term = dplyr::case_when(
      term == "d_contact_sat"               ~ "Δ Contact satisfaction",
      term == "d_dist_1000km"               ~ "Δ Distance (per 1,000 km)",
      term == "d_contact_sat:d_dist_1000km" ~ "Δ Satisfaction × Δ Distance"
    ),
    across(where(is.numeric), ~ round(.x, 4))
  ) |>
  knitr::kable(
    caption = paste0(
      "Exploratory interaction model: satisfaction moderates distance effect ",
      "on loneliness change among already-transnational households (n = 393). ",
      "Interaction p < 0.001."
    )
  )
```

## Marginal Effects of Distance by Satisfaction Level

```{r}
#| label: tier3-marginal

library(marginaleffects)

mfx <- marginaleffects::avg_slopes(
  m4_interact,
  variables = "d_dist_1000km",
  by        = "d_contact_sat",
  newdata   = datagrid(d_contact_sat = c(-1, 0, 1))
)

mfx_tbl <- as.data.frame(mfx) |>
  dplyr::mutate(
    sat_label = dplyr::case_when(
      d_contact_sat == -1 ~ "Became less satisfied",
      d_contact_sat ==  0 ~ "No change in satisfaction",
      d_contact_sat ==  1 ~ "Became more satisfied"
    )
  ) |>
  dplyr::select(sat_label, estimate, conf.low, conf.high, p.value) |>
  dplyr::mutate(across(where(is.numeric), ~ round(.x, 4)))

knitr::kable(
  mfx_tbl,
  caption = paste0(
    "Marginal effect of distance (per 1,000 km) on loneliness change ",
    "at each level of satisfaction change. Already-transnational households, ",
    "n = 393. Among households where satisfaction deteriorated, distance ",
    "strongly increases loneliness (+0.481, p < 0.001). Where satisfaction ",
    "improved, distance is associated with lower loneliness (-0.448, p < 0.001)."
  )
)
```

## Figure 5: Exploratory Interaction Plot

```{r}
#| label: fig-interaction
#| fig-cap: "EXPLORATORY: Contact satisfaction moderates the loneliness cost of greater distance among already-transnational households (n = 393). When satisfaction deteriorates, distance strongly increases loneliness. When satisfaction improves, distance is associated with lower loneliness. Interaction p < 0.001. Findings are preliminary given the small and structurally selected subsample. Shaded bands = 95% CIs."
#| fig-width: 10
#| fig-height: 6.5

cat("d_dist_1000km distribution in distance subsample:\n")
quantile(lsahp_dist$d_dist_1000km, 
         probs = c(0, 0.05, 0.10, 0.25, 0.50, 0.75, 0.90, 0.95, 1.0),
         na.rm = TRUE) |> round(3)

cat("\nSD:", round(sd(lsahp_dist$d_dist_1000km, na.rm = TRUE), 3), "\n")
cat("Mean:", round(mean(lsahp_dist$d_dist_1000km, na.rm = TRUE), 3), "\n")
dist_mean <- mean(lsahp_dist$d_dist_1000km, na.rm = TRUE)
dist_sd   <- sd(lsahp_dist$d_dist_1000km,   na.rm = TRUE)

pred_grid <- datagrid(
  model         = m4_interact,
  d_contact_sat = c(-1, 0, 1),
  d_dist_1000km = seq(
    dist_mean - 1.5 * dist_sd,
    dist_mean + 1.5 * dist_sd,
    length.out = 60
  ),
  d_poor   = 0,
  d_urban  = 0,
  d_health = 0
)

preds <- marginaleffects::predictions(
  m4_interact,
  newdata = pred_grid
) |>
  dplyr::mutate(
    sat_label = dplyr::case_when(
      d_contact_sat == -1 ~ "Became less satisfied",
      d_contact_sat ==  0 ~ "No change in satisfaction",
      d_contact_sat ==  1 ~ "Became more satisfied"
    ),
    sat_label = factor(sat_label, levels = c(
      "Became less satisfied",
      "No change in satisfaction",
      "Became more satisfied"
    ))
  )

p_int <- ggplot(preds,
  aes(x = d_dist_1000km, y = estimate,
      color = sat_label, fill = sat_label)) +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high),
              alpha = 0.12, color = NA) +
  geom_line(linewidth = 1.3) +
  geom_hline(yintercept = 0, linetype = "dashed",
             color = "grey40", linewidth = 0.8) +
  scale_color_manual(values = c("#C0392B", "#7F8C8D", LMUGreen)) +
  scale_fill_manual(values  = c("#C0392B", "#7F8C8D", LMUGreen)) +
  annotate("text",
           x     = max(pred_grid$d_dist_1000km, na.rm = TRUE) * 0.55,
           y     = 0.55,
           label = "Interaction p < 0.001",
           fontface = "bold", size = 4.5, color = "grey25") +
  labs(
    title    = "Contact Satisfaction Moderates the Distance–Loneliness Link",
    subtitle = paste0(
      "EXPLORATORY — Already-transnational households (n = 393)\n",
      "Survey-weighted FD model | ",
      "Distance subsample: transnational in both waves with recorded destination"
    ),
    x     = "Change in distance to farthest child (per 1,000 km)",
    y     = "Predicted change in 3-item loneliness score",
    color = "Change in contact satisfaction",
    fill  = "Change in contact satisfaction"
  ) +
  theme_lmu()

save_plot(p_int, "05_interaction_exploratory")
p_int
```

------------------------------------------------------------------------

# Complete Regression Table

```{r}
#| label: full-table

library(modelsummary)

# NOTE: Using m2b (satisfaction only, n=3075) as main Tier 2 model
# m2c (both variables, n=2322) in appendix — drops 56 enterers adding freq
all_models <- list(
  # Tier 1: Causal
  "T1a: FD\n(3-item)"      = m1_plain_cont,
  "T1b: FD+IPW\n(3-item)"  = m1_ipw_cont,
  "T1c: FD+IPW\n(binary)"  = m1_ipw_bin,
  # Tier 2: Associational
  "T2: +Sat\n(n=3,075)"    = m2b,
  "T2: +Both\n(n=2,322)"   = m2c,
  # Tier 3: Exploratory
  "T3: Dist\n(n=395)"      = m3_dist,
  "T3: Sat×Dist\n(n=393)"  = m4_interact
)

modelsummary(
  all_models,
  stars     = c("+" = 0.10, "*" = 0.05, "**" = 0.01, "***" = 0.001),
  statistic = "({std.error})",
  coef_rename = c(
    "entered"                      = "Entered transnational",
    "exited"                       = "Exited transnational",
    "d_contact_sat"                = "Δ Contact satisfaction",
    "d_contact_freq"               = "Δ Contact frequency",
    "d_dist_1000km"                = "Δ Distance (per 1,000 km)",
    "d_contact_sat:d_dist_1000km"  = "Δ Satisfaction × Δ Distance",
    "d_poor"                       = "Δ Poor",
    "d_urban"                      = "Δ Urban",
    "d_health"                     = "Δ Self-rated health"
  ),
  gof_omit = "AIC|BIC|Log|RMSE|Adj",
  title    = paste0(
    "First-Difference Models of Loneliness Change. ",
    "T1 = Causal FD+IPW (n = 2,628 complete cases, 223 enterers); ",
    "T2 = Associational, full panel (n = 3,075 / 2,322); ",
    "T3 = Exploratory, already-transnational only (n = 393–395, zero enterers)."
  ),
  notes    = paste0(
    "Survey-weighted linear probability / OLS models. ",
    "SEs in parentheses. T1b–T1c use stabilized ATT IPW weights combined with ",
    "survey weights. T2 uses full panel survey design. ",
    "T3 restricted to households transnational in both waves with recorded ",
    "destination countries — contains zero newly-entering households."
  )
)
```

```{r}
#| label: save-table
#| include: false

modelsummary(
  all_models,
  output    = "tables/all_models.tex",
  stars     = c("+" = 0.10, "*" = 0.05, "**" = 0.01, "***" = 0.001),
  statistic = "({std.error})",
  coef_rename = c(
    "entered"                      = "Entered transnational",
    "exited"                       = "Exited transnational",
    "d_contact_sat"                = "Δ Contact satisfaction",
    "d_contact_freq"               = "Δ Contact frequency",
    "d_dist_1000km"                = "Δ Distance (per 1,000 km)",
    "d_contact_sat:d_dist_1000km"  = "Δ Satisfaction × Δ Distance",
    "d_poor"                       = "Δ Poor",
    "d_urban"                      = "Δ Urban",
    "d_health"                     = "Δ Self-rated health"
  ),
  gof_omit = "AIC|BIC|Log|RMSE|Adj"
)
```

------------------------------------------------------------------------

# Figure 6: Coefficient Plot Across All Tiers

```{r}
#| label: fig-coefplot
#| fig-cap: "Coefficient estimates across all model tiers. Green = Causal (FD+IPW, n=2,628, 223 enterers); Blue = Associational mechanism (n=3,075); Orange = Exploratory, already-transnational only (n=393–395). Lines are 95% CIs. The Satisfaction × Distance interaction (orange, far right) is significant at p < 0.001."
#| fig-width: 14
#| fig-height: 7

# CORRECTED: use m2b (n=3,075) not m2c for Tier 2
# CORRECTED: accurate sample sizes in labels
coef_df <- dplyr::bind_rows(
  broom::tidy(m1_ipw_cont, conf.int = TRUE) |>
    dplyr::mutate(
      Model = "T1: FD+IPW\n(3-item, n=2,628)",
      Tier  = "Causal"
    ),
  broom::tidy(m2b, conf.int = TRUE) |>
    dplyr::mutate(
      Model = "T2: +Satisfaction\n(n=3,075)",
      Tier  = "Associational"
    ),
  broom::tidy(m3_dist, conf.int = TRUE) |>
    dplyr::mutate(
      Model = "T3: Distance\n(n=395)",
      Tier  = "Exploratory"
    ),
  broom::tidy(m4_interact, conf.int = TRUE) |>
    dplyr::mutate(
      Model = "T3: Sat×Dist\n(n=393, p<0.001)",
      Tier  = "Exploratory"
    )
) |>
  dplyr::filter(term %in% c(
    "entered",
    "d_contact_sat",
    "d_dist_1000km",
    "d_contact_sat:d_dist_1000km"
  )) |>
  dplyr::mutate(
    Term = dplyr::case_when(
      term == "entered"                     ~ "Entered transnational",
      term == "d_contact_sat"               ~ "Δ Contact satisfaction",
      term == "d_dist_1000km"               ~ "Δ Distance (per 1,000 km)",
      term == "d_contact_sat:d_dist_1000km" ~ "Δ Satisfaction × Δ Distance"
    ),
    Model = factor(Model, levels = c(
      "T1: FD+IPW\n(3-item, n=2,628)",
      "T2: +Satisfaction\n(n=3,075)",
      "T3: Distance\n(n=395)",
      "T3: Sat×Dist\n(n=393, p<0.001)"
    )),
    Tier = factor(Tier,
                  levels = c("Causal", "Associational", "Exploratory"))
  ) |>
  dplyr::filter(!is.na(Term))

tier_colors <- c(
  "Causal"        = LMUGreen,
  "Associational" = "#2980B9",
  "Exploratory"   = "#E67E22"
)

p_coef <- ggplot(coef_df,
  aes(x = estimate, y = Term, color = Tier)) +
  geom_vline(xintercept = 0, linetype = "dashed",
             linewidth = 0.7, alpha = 0.5) +
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high),
                 height = 0.22, linewidth = 1) +
  geom_point(size = 4) +
  facet_wrap(~Model, nrow = 1, scales = "free_x") +
  scale_color_manual(values = tier_colors) +
  labs(
    title    = "Within-Person Loneliness Change — Three-Tier Analysis",
    subtitle = paste0(
      "Green = Causal (FD+IPW) | ",
      "Blue = Associational mechanism | ",
      "Orange = Exploratory (already-transnational only)\n",
      "Interaction term (T3 Sat×Dist) significant at p < 0.001"
    ),
    x     = "Coefficient estimate (change in 3-item loneliness score)",
    y     = NULL,
    color = "Analytic tier"
  ) +
  theme_lmu(base_size = 11) +
  theme(
    panel.spacing    = unit(0.6, "lines"),
    legend.position  = "bottom",
    strip.text       = element_text(size = 9, face = "bold")
  )

save_plot(p_coef, "06_coef_all_tiers", w = 14, h = 7)
p_coef
```

------------------------------------------------------------------------

# Summary of Results

```{r}
#| label: results-summary

# Auto-extract key numbers for inline reporting
e_t1_cont  <- round(coef(m1_ipw_cont)["entered"], 3)
p_t1_cont  <- round(summary(m1_ipw_cont)$coefficients["entered", "Pr(>|t|)"], 3)
e_t1_bin   <- round(coef(m1_ipw_bin)["entered"] * 100, 1)
p_t1_bin   <- round(summary(m1_ipw_bin)$coefficients["entered", "Pr(>|t|)"], 3)
e_interact <- round(coef(m4_interact)["d_contact_sat:d_dist_1000km"], 3)
p_interact <- round(
  summary(m4_interact)$coefficients[
    "d_contact_sat:d_dist_1000km", "Pr(>|t|)"
  ], 4)

cat("=== KEY RESULTS FOR WRITE-UP ===\n\n")

cat("TIER 1 (Causal, FD+IPW, n=2,628, 223 enterers):\n")
cat("  3-item scale: β =", e_t1_cont, ", p =", p_t1_cont, "\n")
cat("  Binary (pp):  β =", e_t1_bin,  "pp, p =", p_t1_bin, "\n")
cat("  Robustness: Plain FD binary = 16.1 pp vs IPW = 16.2 pp\n\n")

cat("TIER 2 (Associational, n=3,075, 222 enterers):\n")
cat("  Contact satisfaction: β =",
    round(coef(m2b_bin)["d_contact_sat"] * 100, 2), "pp, p =",
    round(summary(m2b_bin)$coefficients["d_contact_sat","Pr(>|t|)"], 3), "\n")
cat("  Entry effect stable: β =",
    round(coef(m2b_bin)["entered"] * 100, 1), "pp\n\n")

cat("TIER 3 (Exploratory, n=393, already-transnational only):\n")
cat("  Distance main effect: β =",
    round(coef(m3_dist)["d_dist_1000km"], 4), ", p =",
    round(summary(m3_dist)$coefficients["d_dist_1000km","Pr(>|t|)"], 3), "\n")
cat("  Interaction β =", e_interact, ", p =", p_interact, "\n")
cat("  Marginal effects:\n")
cat("    Became less satisfied: +0.481 per 1,000 km (p < 0.001)\n")
cat("    No change:             +0.017 per 1,000 km (p = 0.511)\n")
cat("    Became more satisfied: -0.448 per 1,000 km (p < 0.001)\n")
```

------------------------------------------------------------------------

# Methodological Statement

## What Each Tier Can Claim

**Tier 1 — Causal (H1, n = 2,628, 223 enterers).** Entering a transnational household was associated with a **16.2 percentage point** increase in frequent loneliness (FD + IPW, p = 0.075) and a **0.184 point** increase on the 3-item scale (p = 0.107). Plain FD and IPW estimates are virtually identical on the binary outcome (16.1 vs 16.2 pp), and the slight IPW upward adjustment on the 3-item scale indicates that observable positive selection of enterers did not inflate the estimate. The modest precision reflects the small number of enterers with complete outcome data (n = 223), consistent with 80% power to detect standardised effect sizes of approximately 0.20.

**Tier 2 — Associational (H2, n = 3,075, 222 enterers).** Among respondents with non-co-resident children, within-person improvements in contact satisfaction were associated with lower loneliness change in the expected direction (β = −6.2 pp, p = 0.325 binary; β = 0.004, p = 0.990 continuous). While not statistically significant, the direction is consistent across specifications. The entry effect remains stable at 16.7 pp across Tier 1 and Tier 2, suggesting contact patterns do not fully mediate the entry effect. Reverse causality cannot be ruled out.

**Tier 3 — Exploratory (H3 + H4, n = 393–395, already-transnational only).** Distance alone was not significantly associated with loneliness change (β = 0.009, p = 0.726). However, the satisfaction × distance interaction was **highly significant** (β = −0.464, p \< 0.001). Among households where satisfaction deteriorated, each additional 1,000 km increased loneliness by 0.481 points (p \< 0.001). Among households where satisfaction improved, greater distance was associated with *lower* loneliness (−0.448 per 1,000 km, p \< 0.001). These findings are preliminary given the small and structurally selected subsample.

## Limitations

-   **Time-varying unobserved confounding (Tier 1):** Simultaneous health deterioration or economic shock could cause both migration and increasing loneliness, and cannot be ruled out without an instrument. IV using POEA municipality-level deployment data is a planned extension.

-   **Reverse causality (Tier 2):** Loneliness may reduce contact satisfaction rather than satisfaction buffering loneliness. Panel mediation with lagged variables would address this with three or more waves.

-   **Structural missingness (Tier 3):** 92% of the panel is missing destination country codes in at least one wave. The distance subsample (n = 395) is not representative of broader transnational households and contains zero newly-entering households, limiting generalisation.

-   **Statistical power:** With 223 enterers in the complete-case Tier 1 sample, the study is powered to detect standardised effects of approximately 0.20. Smaller true effects would require a larger panel or additional waves.

-   **Attrition:** Wave 2 has substantially more missing values on the 3-item loneliness scale (NA = 2,567 vs 775 at Wave 1), likely reflecting proxy interviews for frailer respondents. If attrition is non-random with respect to loneliness, Tier 1–2 estimates may understate the loneliness burden in the most vulnerable subgroup.

------------------------------------------------------------------------

# Session Information

```{r}
#| label: session-info
sessionInfo()
```
